---
title: "Analysis of Stream Annual Temperature Cycles"
author: "Daniel Philippus"
date: "`r format(Sys.Date(), format='%b %d, %Y)')`"
output: pdf_document
---

# Setup

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE)
library(tidyverse)
library(cowplot)
source("functions.R")
source("util.R")
```

# Retrieve USGS Data

There are 67,706 gages with temperature data in US states.

## Identify Gages

```{r eval=FALSE}
all.tg <- map_dfr(state.abb, list.gages) %>%
  as_tibble %>%
  select(site_no,
         lat=dec_lat_va,
         lon=dec_long_va)
```
```{r eval=FALSE}
all.tg$tslen <- validate.gages(all.tg$site_no, log=T)
write_csv(all.tg, "Data/GageList.txt")
```

## Retrieve Data

### Gage List

886 gages have at least 10 years of data.  About 2900 have at least one year.
1500 have at least 5 years.  Maximum duration is 48 years.

```{r}
readgl <- TRUE

tg <- (if (readgl) {
  read_csv("Data/GageList.txt", col_types = "cddi")
  } else all.tg) %>%
  filter(tslen >= 365)

summary(tg)
```

#### Generate GEE Gage List

For retrieving data about gages.  Format: `[[lon, lat], "id"],`

```{r eval=FALSE}
tg %>%
  mutate(
    r1 = paste0("[[", lon),
    r2 = paste0(lat, "]"),
    r3 = paste0('"', site_no, '"]')
  ) %>%
  select(r1:r3) %>%
  write_delim("EE/AllPts.txt", col_names = F, delim=",", eol=",\n", quote = "none", escape="none")
```



### Gage Properties

Of the ~1500 gages with 5 years' data, 1250 have non-NA information.

```{r}
gage.info <- (if (file.exists("Data/GageInfo.txt")) {
  read_csv("Data/GageInfo.txt", col_types = "cdcd")
} else {
  gage.info <- drop_na(get.gageinfo(tg$site_no))
  write_csv(gage.info, "Data/GageInfo.txt")
  gage.info
}) %>%
  group_by(id) %>%
  slice_head(n=1) %>%
  ungroup()

summary(gage.info)
```

### Temperature Records

Gages with available info and at least 5 years' data provide about 6 million rows,
with coverage from WY1965-present.  There are 1,234 gages with successfully 
retrieved data, making a mean of ~5000 days (14 years) per gage.

```{r eval=FALSE}
start <- "1950-01-01"
end <- "2023-01-01"

temps <- if (file.exists("Data/GageTemperatures.csv")) {
  read_csv("Data/GageTemperatures.csv", col_types = "cDid")
} else {
  gns <- gage.info$id
  is.done <- map_lgl(gns, ~file.exists(paste0("Data/GageChunks/", ., ".csv")))
  print(paste("Done:", sum(is.done)))
  cat("\n")
  for (gn in gns[!is.done]) {
    cat("\rRunning", gn)
    write_csv(get.usgs(gn, start, end), paste0("Data/GageChunks/", gn, ".csv"))
  }
  temps <- map_dfr(paste0("Data/GageChunks/", gns, ".csv"),
                   ~drop_na(read_csv(., col_types = "cDid")))
  write_csv(temps, "Data/GageTemperatures.csv")
  temps
}

summary(temps)
```

Quality control:

```{r eval=FALSE}
tqc <- temps %>%
  filter(temperature >= 0,
         temperature <= 100)
```


## Full Dataset


```{r eval=FALSE}
data <- left_join(gage.info, tqc, by="id") %>%
  drop_na
write_csv(data, "Data/GageData.csv")
summary(data)
```

## Load Full Dataset

```{r}
data <- read_csv("Data/GageData.csv",
                 col_types = "cdcdDid") %>%
  drop_na
```

```{r}
data %>%
  group_by(id, day) %>%
  summarize(t = mean(temperature)) %>%
  summary
```

```{r}
data %>%
  group_by(id) %>%
  summarize(t = mean(temperature)) %>%
  summary
```

# Gage Timeseries Map

```{r}
plt <- plot.eco(TRUE, FALSE, 4) +
  # scale_fill_viridis_d(option="magma", end=0.9) +
  geom_point(aes(lon, lat, size=tslen/365), data={
    left_join(data, tg, by=c("id"="site_no")) %>%
      group_by(id) %>%
      summarize(lon=first(lon),
                lat=first(lat),
                tslen=length(unique(Date)))
  }, color="black") +
  # scale_color_viridis_c(begin=0.5) +
  theme_std() +
  theme(legend.position = "bottom",
        legend.direction = "vertical") +
  labs(
    x="Longitude",
    y="Latitude",
    size="Timeseries Length (years)",
    fill="EPA Level I Ecoregion"
  )
write.img(plt, "Figures/GageCovMap_ts", height=5)
```

# Overview By Elevation

## Observed Rolling Means By Bucket

```{r}
plt <- bucketize(data, "elev", "m", 10) %>%
  group_by(id) %>%
  mutate(
    temp = temperature / mean(temperature)
  ) %>%
  group_by(BucketName, day) %>%
  summarize(
    temp = median(temp)
  ) %>%
  group_by(BucketName) %>%
  arrange(day) %>%
  mutate(temp = stats::filter(temp, rep(1/30, 30), circular=TRUE)) %>%
  ggplot() +
  aes(day, temp, color=BucketName) +
  geom_line(size=2) +
  scale_color_viridis_d(option="magma", end=0.8) +
  scale_x_continuous(breaks=30*(0:12)) +
  theme_std(size=36) +
  theme(legend.position = "bottom",
        axis.text = element_text(size=30),
        legend.text = element_text(size=30)) +
  labs(x="Day of Year",
       y="Bucket Median Normalized Stream Temperature\n(30-Day Mean)",
       color="Elevation Bucket")

print(plt + theme_std(12))
write.img(plt, "Figures/ObservedElevationBuckets")
# ggsave("Figures/ObservedElevationBuckets.eps", plt, width=6.5, height=4)
```

## Model Data Preparation

```{r}
fitted <- data %>%
  sinu.resfit %>%
  drop_na() %>%
  bucketize("elev", "m", 10) %>%
  group_by(id) %>%
  group_modify(~{
    left_join(.x, make.sints(fit.sins(.x)), by="day")
    })
```

## Fourier Elevation Comparison

When `fixed` is TRUE (one extra variable), there is a very, very slight performance
penalty to the 3-sin model (<0.003 for all buckets and typ. <0.001 median).  When
`fixed` is FALSE (four extra variables), it has a small performance advantage for
all buckets except 185-229 m, but growing with elevation (up to 0.009).  Adding
a third harmonic (so two extra variables) reintroduces a small performance penalty
(0.002-0.003).

```{r warning=FALSE}
fitted <- data %>%
  group_by(id, day, elev) %>%
  summarize(temperature = mean(temperature)) %>%
  ungroup() %>%
  bucketize("elev", "m", 10) %>%
  group_by(id, BucketName) %>%
  arrange(day) %>%
  group_modify(~cbind(
    fit.sins(.x, fixed=FALSE),
    tibble(FR2 = fourier(.x$temperature, 3)$R2)))
```


```{r}
fitted %>%
  drop_na() %>%
  group_by(BucketName) %>%
  summarize(Gap = median(R2 - FR2),
            across(c(R2, FR2),
                   median))
```


## Residual Plot

```{r}
plt <- data %>%
  bucketize("elev", "m", 10) %>%
  group_by(BucketName, day) %>%
  summarize(
    temperature = mean(temperature)
  ) %>%
  group_by(BucketName) %>%
  arrange(day) %>%
  mutate(
    sinusoid = lm(temperature ~ cos((day - 210) * 2 * pi / 365))$fitted.values,
    residual = stats::filter(
      (temperature - sinusoid) / mean(temperature),
      rep(1/31, 31),
      circular=T
    )
  ) %>%
  ggplot(aes(day, residual, color=BucketName)) +
  geom_line(size=2) +
  scale_color_viridis_d(end=0.9) +
  theme_std() +
  theme(legend.position = "bottom") +
  labs(
    x="Day of Year",
    y="Mean Temperature Residual Compared to Sinusoid\n(31-Day Rolling Mean)",
    color="Elevation Bucket"
  )
write.img(plt, "Figures/Residual")
```



## Comparison Plot

This shows the point very well.  Model behavior is nearly indistinguishable at
low elevation, while the anomaly model clearly fits observed behavior better at
high elevation.  However, it is evident that even at lower elevations the anomaly
model does better; it is just a subtle enough effect not to be obvious unless
specifically looked for.

```{r}
fitted <- data %>%
  sinu.resfit %>%
  drop_na() %>%
  bucketize("elev", "m", 10) %>%
  group_by(id) %>%
  group_modify(~{
    left_join(.x, make.sints(fit.sins(.x)), by="day")
    })
```

```{r}
plt <- fitted %>%
  group_by(BucketName, day) %>%
  summarize(
    across(c(ntemp, reftemp, RefMod, actemp),
           ~median(.))
  ) %>%
  group_by(BucketName) %>%
  mutate(actemp = actemp / mean(actemp)) %>%
  arrange(day) %>%
  mutate(
    across(c(ntemp, reftemp, RefMod, actemp),
           ~stats::filter(., rep(1/30, 30), circular=TRUE))
  ) %>%
  rename(Sinusoidal = reftemp,
         `Anomaly Model` = RefMod,
         `Three-Sinusoid Model` = actemp) %>%
  pivot_longer(c(Sinusoidal, `Anomaly Model`, `Three-Sinusoid Model`),
               names_to="Model",
               values_to="Prediction") %>%
  ggplot() +
  aes(day) +
  geom_line(aes(y=Prediction, color=Model), size=1) +
  geom_point(aes(y=ntemp, color="Observed"), size=1) +
  facet_wrap(~BucketName, nrow=2) +
  scale_color_viridis_d() +
  theme_std() +
  labs(
    x="Day of Year",
       y="Bucket Median Normalized Stream Temperature\n(30-Day Mean)",
       color="Model Type"
  )

print(plt + theme_std(12))
write.img(plt, "Figures/ElevationModelComparison")
```

## Model Performance Comparison

```{r}
perf <- fitted %>%
  drop_na() %>%
  group_by(id, day) %>%
  summarize(
    across(c(ntemp, reftemp, actemp),
           mean),
    BucketName = first(BucketName)
  ) %>%
  group_by(id) %>%
  summarize(
    BucketName = first(BucketName),
    across(c(reftemp, actemp),
           ~cor(.x, ntemp)^2),
    Delta = actemp - reftemp
  ) %>%
  ungroup()
```

The anomaly model outperforms the sinusoid model for about 3/4 of gages.  The
typical difference is modest, but it closes about half the small gap to perfect
performance.  Importantly, the anomaly model does much better for some marginally
worse-performing areas (e.g. R2 around 0.9-0.95).

The performance gap grows quite rapidly with elevation buckets.  From 0-60%,
the gap in R2 is under 0.02, growing by around 0.001-0.003 per bucket.  It then
grows quite rapidly up to 90%, maxing out at a gap of almost 0.05.  This mirrors
the visible behavior in the plots.

```{r}
summary(select(perf, reftemp, actemp, Delta))
```
```{r}
perf %>%
  group_by(BucketName) %>%
  summarize(
    across(c(reftemp, starts_with("ac"), starts_with("Delta")),
           median)
  )
```

## Statistically Significant Difference

Test differences between the mean and the sinusoid (`reftemp`) and 3sin (`actemp`)
temperatures.

```{r}
difs <- fitted %>%
  group_by(id) %>%
  mutate(sinusoid = reftemp * mean(temperature)) %>%
  group_by(id, day) %>%
  filter(n() >= 30) %>%
  summarize(
    Mean = mean(temperature),
    SD = sd(temperature),
    across(c(sinusoid, actemp),
           list(
             "zscore" = ~(mean(.x) - Mean) / (SD / sqrt(n())),
             "pval" = ~pnorm(mean(.x), mean=Mean, sd=SD/sqrt(n()), lower.tail=(.x <= Mean))
           ))
  )
```
```{r}
summary(select(ungroup(difs), Mean:actemp_pval))
```
```{r}
difs %>%
  group_by(id) %>%
  summarize(
    across(
      ends_with("_pval"),
      ~sum(.x < 0.05) / n()
    )
  ) %>%
  summary
```



# Three-Sinusoid Fit Testing

Use high elevation to test.

```{r}
uel <- data %>% bucketize("elev", "m", 10) %>% filter(Bucket == 90)
```

We have the usual visibly non-sinusoidal shape.

```{r}
uel.mean <- uel %>%
  group_by(day) %>%
  summarize(
    temperature = mean(temperature, na.rm=T)
  )
plot(uel.mean$day, uel.mean$temperature)
```

```{r}
sinft <- fit.sins(uel.mean)
fitted <- uel.mean %>%
  add.sinusoid.all %>%
  add.resfit %>%
  left_join(make.sints(sinft),
            by="day")
```

The sinusoidal model has the usual shortcomings.

```{r}
plot(fitted$day, fitted$temperature)
lines(fitted$day, fitted$reftemp * mean(fitted$temperature), col="red")
```
The anomaly model, on the other hand, does quite well, as per usual.

```{r}
plot(fitted$day, fitted$temperature)
lines(fitted$day, fitted$RefMod * mean(fitted$temperature), col="red")
```

The three-sinusoid model also does very well and behaves quite similarly.
They trade off doing well or poorly, but the 3S model does better at the maximum
and minimum.

```{r}
plot(fitted$day, fitted$temperature)
lines(fitted$day, fitted$actemp, col="red")
lines(fitted$day, fitted$RefMod * mean(fitted$temperature), col="green")
```

The 3S model is very marginally the best performer, and both solidly outdo the
sinusoid model.

```{r}
cor(fitted$temperature, fitted$reftemp)^2
cor(fitted$temperature, fitted$RefMod)^2
cor(fitted$temperature, fitted$actemp)^2
```

# Demo Plots

Replicate preliminary figures from TE2 seasonality work: modeled anomalies.

```{r}
fits <- data %>%
  bucketize("elev", "m", 20) %>%
  group_by(BucketName, day) %>%
  summarize(temperature = mean(temperature)) %>%
  group_by(BucketName) %>%
  group_modify(~{
    fit <- fit.sins(.x)
    prd <- make.sints(fit)
    tibble(
      day = prd$day,
      actemp = prd$actemp - (fit$Intercept + fit$Amplitude * cos((day - 210) * 2 * pi / 365))
    )
  })
```

```{r}
plt <- ggplot(fits, aes(day, actemp, color=BucketName)) +
  geom_line(size=2) +
  scale_color_viridis_d() +
  theme_std() +
  theme(legend.position = "bottom") +
  labs(
    x="Day of Year",
    y="Predicted Seasonal Temperature Residual",
    color="Elevation\nBucket"
  )
write.img(plt, "Figures/ElevationAnomFitted")
```




# Formally Characterizing Model Differences

## lm Statistical Significance

```{r}
lms <- data %>%
  group_by(id) %>%
  group_modify(~{
    safely(\(x) summary(fit.sins(x, TRUE))$coefficients[,4] %>%
      as.list() %>%
      as_tibble,
      otherwise=tibble(sin1=NA, sin2=NA))(.)$result
  })
```
```{r}
summary(select(lms, -id))
```

```{r}
quantile(lms$sin1, 1:100 / 100, na.rm=T)
```
```{r}
quantile(lms$sin2, 1:100 / 100, na.rm=T)
```

```{r}
lms <- data %>%
  bucketize("elev", "m", 10) %>%
  group_by(BucketName) %>%
  group_modify(~{
    safely(\(x) summary(fit.sins(x, TRUE))$coefficients[,4] %>%
      as.list() %>%
      as_tibble,
      otherwise=tibble(sin1=NA, sin2=NA))(.)$result
  })
```
```{r}
lms
```

## Compare AIC

Use `fit.sins(mod=TRUE, aic=TRUE)` to compute AIC.

```{r}
aicda <- data %>%
  group_by(id) %>%
  group_modify(~fit.sins(.x, TRUE, TRUE)) %>%
  mutate(Delta = Anom - Sin) %>%
  ungroup() %>%
  select(-id)
```
```{r}
summary(aicda)
quantile(aicda$Delta, (1:100)/100, na.rm=T)
```
```{r}
abkt <- data %>%
  bucketize("elev", "m", 10) %>%
  group_by(BucketName) %>%
  group_modify(~fit.sins(.x, TRUE, TRUE)) %>%
  mutate(Delta = Anom - Sin)
abkt
```

```{r}
abkt %>%
  mutate(
    across(Sin:Delta, as.integer)
  ) %>%
  rename(
    `Elevation Bucket`=BucketName,
    `Sinusoidal Model AIC`=Sin,
    `Three-Sine Model AIC`=Anom,
    `AIC Difference (Three-Sine minus Sinusoidal)`=Delta
  ) %>%
  write_delim("AICtab.txt",
              delim="&",
              eol="\\\\\n",
              quote="none")
```
```{r}
plt <- abkt %>%
  rename(
    `Sinusoidal Model`=Sin,
    `Three-Sine Model`=Anom,
    `Difference (Three-Sine minus Sinusoidal)`=Delta
  ) %>%
  pivot_longer(-BucketName) %>%
  ggplot(aes(BucketName, value, fill=name)) +
  geom_col(position="dodge") +
  scale_fill_viridis_d(end=0.8,
                        guide_legend(title="Model",
                                     title.position="left",
                                     direction = "horizontal")) +
  theme_std() +
  theme(legend.text = element_text(size=30)) +
  theme(legend.position = "bottom") +
  labs(
    x = "Elevation Bucket",
    y = "Akaike Information Criterion",
    fill = "Model"
  )
write.img(plt, "Figures/AICBars", height=2.5)
```


## Compare to Fourier Transform

### 3S Model Performance Distribution

```{r}
fitted <- data %>%
  group_by(id) %>%
  filter(length(unique(day)) >= 365) %>%
  group_modify(~fit.sins(.x, fixed=TRUE)) %>%
  ungroup()
```
```{r}
summary(fitted$R2)
summary(fitted$RMSE)
```
### Fourier Transform

```{r message=FALSE}
foured <- data %>%
  group_by(id, day) %>%
  summarize(temperature = mean(temperature)) %>%
  group_by(id) %>%
  filter(length(unique(day)) >= 365) %>%
  summarize(fourier(temperature, 2))

summary(foured)
```

Note that the same amount of data is required when N=4 (real and imaginary part).

For a median R2 of 0.98-0.99 (IQR 0.97-0.99) and RMSE of 0.78 C (0.58-1.07), about
two terms are required (four numbers).  That's in comparison to fitting all four
dates (eight numbers).  With `fixed=TRUE` (five numbers), it is 0.98 (0.96-0.99)
and 0.87 (0.65-1.19), which still requires N=2 (five numbers vs four).

An alternative approach is asymmetrical, since the first Fourier term doesn't
wrap around.  However, this doesn't affect the behavior much.


# Using the Model - For Science!

```{r}
data <- read_csv("Data/GageData.csv",
                 col_types = "cdcdDid") %>%
  drop_na
```

## Geographic Trends

### Moving Window

Compute overall seasonality for all rivers within 2 degrees centered on each
integer coordinate pair.  Plot each trend.

```{r}
geo <- filter(left_join(data, tg, by=c("id"="site_no")), lon >= -130) %>%
  drop_na
```

```{r warning=FALSE}
lons <- (-130):(-65)
lats <- 25:50
width <- 2

geossn <- map_dfr(lons,
                  function(x) {
                    map_dfr(lats,
                            function(y) {
                              raw <- filter(geo,
                                            lon >= x - width/2,
                                            lon < x + width/2,
                                            lat <= y + width/2,
                                            lat > y - width/2)
                              if (nrow(raw) > 0) {
                                cbind(
                                  tibble(
                                    lon = x,
                                    lat = y,
                                    temperature = mean(raw$temperature)
                                  ),
                                  fit.sins(raw)
                                ) %>%
                                  normalize.sins()
                              } else tibble()
                            })
                  })
write_csv(geossn, "Data/GeoWindow.csv")
```


Some variables range quite substantially by location.  For example, the spring
low and winter high can range from negligible to a large fraction of an amplitude.

```{r}
geossn <- read_csv("Data/GeoWindow.csv", show_col_types=F)
summary(geossn)
```

In these maps:

- Temperatures are high south, low, and near the Gulf.
- Amplitudes are high north, inland, and in Maine.
- The autumn/winter anomaly is large at high elevation and in the northern Great Plains.
The autumn low hits early where amplitude is high, and the winter anomaly hits early
near cold water.  The anomaly can reach up to 20-40% of mean temperatures, which is
a much larger proportion of mean temperatures in the fall/winter.
- The spring/summer anomaly is strongest in the high-elevation west and Maine.
 The spring low hits early around the Great Lakes, and the summer high hits early
 inland.  The anomaly can likewise reach 20-40% of mean temperatures.

```{r}
geossn <- select(geossn, -any_of(c("R2", "RMSE", "temperature")))
states <- map_data("state")
for (nm in names(geossn)[-(1:2)]) {
  plt <- ggplot(geossn,
                aes(lon, lat, fill=.data[[nm]])) +
    geom_polygon(aes(long, lat, group=group), data=states, color="black",
                 fill=NA, size=2) +
    geom_tile(alpha=0.9) +
    scale_fill_viridis_c(
      guide = guide_colorbar(barwidth=unit(400, "points"),
                                               barheight=unit(50, "points"))
    ) +
    theme_std() +
    theme(legend.position = "bottom") +
    labs(
      x="Longitude",
      y="Latitude",
      fill=paste0("Window\n", renamer[nm])
    )
  write.img(plt, paste0("Figures/MovingWindowSeasons/", nm))
}

plt <- geossn %>%
  mutate(
    across(Intercept:SpringSummer,
           ~(.x - mean(.x, na.rm=T))/sd(.x, na.rm=T))
  ) %>%
  pivot_longer(Intercept:SpringSummer,
               names_to="Aspect",
               values_to="Value") %>%
  ggplot(aes(lon, lat, fill=Value)) +
  geom_polygon(aes(long, lat, group=group), data=states, color="black",
                 fill=NA, size=2) +
  geom_tile(alpha=0.9) +
  scale_fill_viridis_c(limits=c(-3, 3),
                       guide = guide_colorbar(barwidth=unit(400, "points"),
                                               barheight=unit(50, "points")))+
  facet_wrap(~Aspect, nrow=2, labeller = as_labeller(renamer)) +
  theme_std() +
  theme(legend.position = "bottom") +
  labs(
    x="Longitude",
    y="Latitude",
    fill="Normalized\nSeasonality\nTerm"
  )
write.img(plt, "Figures/MovingWindowSeasons/All")
```

### Timing Trends

The identified timing trends are difficult to read in text, so we make some plots
of them.  Trends to plot:

- The winter-high is more or less binary between early and late, occurring early
in the Pacific Northwest, northern Rockies, upper Midwest, and far Northeast, and
late elsewhere.  This can be loosely captured by a division around 42 N, though
there are several exceptions.
- The summer-high has three discrete regions.  It is early in the Mountain West,
Great Plains, and Upper Midwest; middling in the Pafific Northwest; and late in the
East and Southeast (again, with exceptions).  The main dividing lines are two
diagonals: 30/-105 to 42/-85 (East/Central) and 40/-123 to 50/-110 (Central/PNW).
- The spring-low divides into East (late, east of -105 and south of 42); Northern
East (early, east of -105 and north of 42); and West (middling, west of -105).
- The autumn-low more or less has an early zone in the north except for the PNW
(-115 eastwards, north of 40) and is middling to late elsewhere.

```{r}
regions <- geossn %>%
  pivot_longer(ends_with("Day"),
               names_to="Which",
               values_to="Day",
               names_pattern = "([A-Za-z]+)Day") %>%
  mutate(
    Which = ifelse(Which == "Fall", "Autumn", Which),
    Zone = case_match(
      Which,
      "Spring" ~ case_when(
        lon > -105 & lat < 42 ~ "East",
        lon > -105 & lat >= 42 ~ "Northern East",
        .default = "West"
      ),
      "Summer" ~ case_when(
        # East/Central line: +12 over +20 -> y = 0.6x + 93
        lat < 0.6 * lon + 93 ~ "Southeast",
        # Central/PNW: +10 over 13 -> y = 10/13x + (50 + 10/13 * 110)
        lat > 10 / 13 * lon + 50 + 10 / 13 * 110 ~ "Northwest",
        .default = "Central"
      ),
      "Autumn" ~ ifelse(lon > -115 &
                          lat > 40, "North", "South/Northwest"),
      "Winter" ~ ifelse(lat > 42, "North", "South")
    )
  )
```

Combined regions:

- Upper Midwest = Autumn North/Winter North/Spring Northern East/Summer Central
- Northeast = AN/WN/SNE/Summer Southeast
- Northern Rockies = AN/WN/Spring West/Summer Central
- Tiny slice of Northern Rockies = AN/WN/SW/Summer Northwest
- Mid-latitude East = four blocks: AN/Winter South/Spring (Northern East/East)/Summer (Central/Southeast)
- Mid-latitude Rockies = AN/WS/SW/SC
- PNW = Autumn Northwest/Winter North/Spring West/Summer (Central/Northwest)
- Southeast = Autumn South/Winter South/Spring East/Summer Southeast
- Central South = AS/WS/SE/Summer Central
- Southwest = AS/WS/Spring West/Summer (Central/Northwest)

Proposed relumping:

```{r}
regions <- geossn %>%
  pivot_longer(ends_with("Day"),
               names_to="Which",
               values_to="Day",
               names_pattern = "([A-Za-z]+)Day") %>%
  mutate(
    Which = ifelse(Which == "Fall", "Autumn", Which),
    Zone = case_when(
      # Midwestern and Eastern Zones
      lon > -105 ~ case_when(
        lat >= 0.6 * lon + 93 ~ case_when(
          lon > -105 & lat > 40 ~ "Northern Midwest",
          .default = "Southern Midwest"
        ),
        lat > 40 ~ "Northeast",
        .default = "Southeast"
      ),
      # Northern West
      lat > 42 ~ case_when(
        lon > -115 ~ "Northern Rocky Mountains",
        .default = "Pacific Northwest"
      ),
      .default = case_when(
        lat > 40 & lon > -115 ~ "Central Rocky Mountains",
        .default = "Southwest"
      )
    )
  )
```


```{r}
reggrp <- regions %>%
  group_by(lon, lat) %>%
  # summarize(AllReg = paste(unique(paste(Which, Zone)), collapse=" | "))
  summarize(AllReg = first(Zone))

plt <- ggplot(reggrp, aes(lon, lat, fill=AllReg)) +
  scale_fill_viridis_d() +
  geom_tile() +
  geom_polygon(aes(long, lat, group=group), data=states, color="black",
                 fill=NA, size=2) +
  theme_std() +
  labs(
    x="Longitude",
    y="Latitude",
    fill="Seasonality\nRegion"
  )
write.img(plt, "Figures/MovingWindowSeasons/SeasonalityRegions")
```


```{r}
plt <- regions %>%
  ggplot(aes(Zone, as.Date(paste("2000", Day), format="%Y %j"),
             color=Zone)) +
  geom_boxplot() +
  scale_color_viridis_d() +
  coord_flip() +
  scale_y_date(labels=scales::date_format("%b"),
               date_breaks="1 month") +
  facet_wrap(~Which, scales="free_y", ncol=1) +
  labs(x="Timing Zone", y="Anomaly Peak Day") +
  theme_std() +
  guides(color="none")
write.img(plt, "Figures/MovingWindowSeasons/RegionDateBoxes")
```

#### Alternative Trends: Clusters

With 5 clusters, there are:

1. Mountain
2. Boundary of warm
3. Cold continental
4. Warm
5. Cold and wet

(2) is sort of redundant with (3), which suggests that 4 clusters might be better.

With 4, there are:

1. Mountain (+NY): cold, large spring/summer anomaly, highly variable summer 
high, early winter high
2. Continental (+some coastal): average on everything
3. Warm: low anomaly coefficients, warm, low amplitude
4. Cold north: cold, large autumn/winter anomaly, largest amplitude, early seasons

```{r}
nona <- drop_na(geossn)
normed <- nona %>%
  mutate(across(Intercept:SpringSummer, ~(.x - mean(.x))/sd(.x))) %>%
  select(Intercept:SpringSummer)

tree <- hclust(dist(normed), "ward.D")
nona$Cluster <- cutree(tree, 4)
```

```{r}
cluster.names <- c(
  "Mountain",
  "Continental",
  "Warm",
  "Cold North"
)

plt_map <- ggplot(nona, aes(lon, lat, fill=cluster.names[Cluster])) +
  geom_tile() +
  geom_polygon(aes(long, lat, group=group), data=states, color="black",
                 fill=NA, size=2) +
  scale_fill_viridis_d(option="plasma", end=0.8) +
  theme_std() +
  theme(legend.position = "bottom") +
  labs(
    x="Longitude",
    y="Latitude",
    fill="Seasonality Cluster"
  )
write.img(plt_map, "Figures/MovingWindowSeasons/Clusters")
```

```{r}
plt <- nona %>%
  pivot_longer(Intercept:SpringSummer,
               names_to="Parameter",
               values_to="Value") %>%
  ggplot(aes(cluster.names[Cluster], Value, color=cluster.names[Cluster])) +
  geom_boxplot(size=2) +
  scale_color_viridis_d(option="plasma", end=0.8) +
  facet_wrap(~Parameter, ncol=4, scales="free_y",
             labeller = as_labeller(renamer),
             strip.position = "left") +
  labs(y=NULL, x="Cluster") +
  theme_vert() +
  guides(color="none")
write.img(plt, "Figures/MovingWindowSeasons/ClusterBoxes")
```
```{r}
plt_box <- nona %>%
  pivot_longer(ends_with("Day"),
               names_to="Parameter",
               values_to="Value") %>%
  mutate(Parameter = factor(Parameter,
                            levels = c(
                              "WinterDay", "SpringDay", "SummerDay", "FallDay"
                            ))) %>%
  ggplot(aes(cluster.names[Cluster], Value, color=cluster.names[Cluster])) +
  geom_boxplot(size=2) +
  scale_y_continuous(limits = \(lims) {
    l <- lims[1]
    h <- lims[2]
    c(l, max(h, l + 70))
  },
  breaks = \(lims) {
    minv <- floor(lims[1] / 20)
    maxv <- floor(lims[2] / 20)
    20 * (minv:maxv)
  }) +
  scale_color_viridis_d(option="plasma", end=0.8) +
  facet_wrap(~Parameter, ncol=3, scales="free_y",
             labeller = as_labeller(renamer),
             strip.position = "left") +
  labs(y=NULL, x="Cluster") +
  theme_vert() +
  guides(color="none")
write.img(plt_box, "Figures/MovingWindowSeasons/ClusterBoxDates")
```

```{r}
plt_map <- plt_map + theme_std(32)
# aligned <- align_plots(plt_box, plt_map, axis="br")
aligned <- ggdraw(plt_box) + draw_plot(plt_map, x=0.4, width=0.6, height=0.4)
write.img(aligned, "Figures/MovingWindowSeasons/ClusterWithInset")
```


### Elevation (100 m)

```{r}
width <- 100

elev <- data %>%
  group_by(elev = width * floor(elev / width), id) %>%
  group_modify(~normalize.sins(fit.sins(.))) %>%
  group_by(elev) %>%
  summarize(
    across(-id,
           list(
             "median" = median,
             "outliers" = ~sum(abs((.x - mean(.x)) / sd(.x)) >= 3)
           )),
    outliers = Intercept_outliers | Amplitude_outliers | FallWinter_outliers | SpringSummer_outliers,
    N = n()
  ) %>%
  rename_with(\(x) str_split(x, "_", simplify=T)[,1], ends_with("_median"))
  # group_modify(~mutate(
  #   normalize.sins(fit.sins(.)), N = length(unique(.$id))
  # ))
```

```{r}
ol.elevs <- c(
                2100,
                2500,
                2600,
                2800)
```

```{r}
olel.names <- c(
  "2100" = "Yellowstone and Eastern Sierra",
  "2500" = "Northern Colorado Rockies",
  "2600" = "Northern/Western Colorado Rockies and Northwestern Wyoming",
  "2800" = "Western/Southwestern Colorado Rockies"
)
```


```{r}
olel <- 2800

left_join(filter(data, elev >= olel, elev < olel + 100),
          tg, by=c("id"="site_no")) %>%
  ggplot() +
  geom_polygon(aes(long, lat, group=group), data=states, fill=NA, color="black") +
  geom_point(aes(lon, lat))
```


```{r}
outliers <- elev %>%
  pivot_longer(c(Intercept, Amplitude, FallWinter, SpringSummer)) %>%
  filter(
    elev %in% ol.elevs
  ) %>%
  mutate(Which = olel.names[as.character(elev)])
```


```{r}
plt <- elev %>%
  pivot_longer(c(Intercept, Amplitude, FallWinter, SpringSummer)) %>%
  ggplot(aes(elev, value, shape=N < 10, group=1)) +
  geom_point(size=4) +
  geom_point(aes(elev, value, shape=N < 10, color=Which), data=outliers, size=8) +
  scale_color_viridis_d(end=0.9, guide=guide_legend(nrow=3,
                                                    title.position = "top",
                                                    override.aes = list(shape=17))) +
  geom_smooth(size=2, method="lm", se=T) +
  facet_wrap(~name, scales="free_y", labeller = as_labeller(renamer),
             strip.position = "left", nrow=2) +
  theme_std() +
  theme(legend.position = "bottom") +
  labs(
    x="Elevation (m)",
    y=NULL,
    shape="Few Rivers (<10)",
    color="Outlier Contents"
  )
write.img(plt, "Figures/ElevationTrends")
```




# Case Study Validation

Based on strength/weakness of spsu and aw anomalies, divergences from sinusoid
should be obvious for western Montana and northern Maine, but negligible for Florida.
Make plots of actual cycle vs sinusoid for:

- Western Montana = -115 to -110; 45+
- Northern Maine = -70+; 45+
- Florida = -87+; 30-

```{r}
frn <- function(lons, lats, name, dat=geo) {
  filter(dat,
         lon >= lons[1], lon <= lons[2],
         lat >= lats[1], lat <= lats[2]) %>%
    mutate(Region = name)
}

regions <- rbind(
  frn(c(-115, -110), c(45, 99), "Western Montana"),
  frn(c(-70, 0), c(45, 99), "Northern Maine"),
  frn(c(-87, 0), c(0, 30), "Florida")
)
```

```{r}
rfit <- regions %>%
  group_by(Region, id, day) %>%
  summarize(temperature = mean(temperature)) %>%
  group_by(Region) %>%
  mutate(
    Cos = cos((day - 210) * 2 * pi / 365),
    Fit = lm(temperature ~ Cos)$fitted.values,
    N = length(unique(id))
  ) %>%
  group_by(Region, day) %>%
  mutate(Mean = mean(temperature)) %>%
  ungroup() %>%
  mutate(Region = paste0(Region, " (n=", N, ")"))
```

```{r}
plt <- ggplot(rfit, aes(day)) +
  geom_point(aes(y=temperature, color="Observed"), size=2, alpha=0.1,
             stroke=NA) +
  geom_line(aes(y=Fit, color="Sinusoid"), size=4) +
  geom_line(aes(y=Mean, color="Mean"), size=2) +
  facet_wrap(~Region, nrow=1) +
  scale_color_viridis_d() +
  theme_std() +
  theme(legend.position = "bottom",
        legend.text = element_text(size=30),
        strip.text = element_text(size=30)) +
  labs(
    x="Day of Year",
    y="Region Mean Temperature",
    color=NULL
  )
write.img(plt, "Figures/GeoValidation_c", height=3)
```

# Temporal Cross-Validation

```{r}
fitted <- data %>%
  sinu.resfit %>%
  drop_na() %>%
  bucketize("elev", "m", 10) %>%
  group_by(id) %>%
  group_modify(~left_join(.x, make.sints(fit.sins(.x)), by="day"))
```

```{r}
cv <- data %>%
  mutate(Year = as.integer(format(Date, "%Y"))) %>%
  group_by(id) %>%
  group_modify(~{
    yrs <- unique(.x$Year)
    cvdat <- left_join(.x,
                      tibble(Year = yrs,
                             Index = sample(1:5, length(yrs), replace=TRUE)),
                      by="Year")
    map_dfr(1:5,
            function(ix) {
              tr <- filter(cvdat, Index != ix)
              te <- filter(cvdat, Index == ix) %>% select(-Index)
              model <- fit.sins(tr)
              days <- 1:366
              if (nrow(drop_na(model)) > 0) {
                modout <- tibble(day=days,
                                 Sinusoid=model$Intercept + model$Amplitude * cos((days - 210) * 2 * pi / 365),
                                 Anomaly=make.sints(model)$actemp)
                left_join(te,
                          modout,
                          by="day")
              } else {
                tibble()
              }
            })
  })
```
```{r}
perf <- cv %>%
  drop_na() %>%
  group_by(id) %>%
  summarize(
    across(c(da, elev), first),
    across(c(Sinusoid, Anomaly),
    ~cor(.x, temperature)^2
  )) %>%
  mutate(Delta = Anomaly - Sinusoid) %>%
  drop_na()

perf %>%
  select(Sinusoid, Anomaly, Delta) %>%
  summary
```

## CV Performance Trends

```{r}
cor(perf$elev, perf$Delta)
cor(perf$da, perf$Delta)
perf %>%
  group_by(Elev500 = 500 * floor(elev/500)) %>%
  summarize(
    across(Sinusoid:Delta,
           median)
  )
```


# Predicting Seasonality Functions

From Isaak et al. 2020: elevation, latitude, riparian canopy density,
reach slope, precipitation, lake prevalence, and dam height

Summarized through PCA.  However, they were using classification categories,
not continuous values.


# Misc

## Gage Counts and Data Coverage

```{r}
plt <- ggplot(tg, aes(x=tslen / 365)) +
  stat_ecdf(size=4) +
  scale_y_continuous(sec.axis = sec_axis(~.*nrow(tg), name="Gage Count")) +
  theme_std() +
  labs(x="Gage Timeseries Length (years)",
       y="Cumulative Probability")
write.img(plt, "Figures/GageECDF", width=4)
```


## Plot of Data Coverage

```{r}
plt <- plot.eco(TRUE) +
  scale_fill_viridis_d(option="magma", end=0.5) +
  geom_point(aes(lon, lat, color=tslen/365, size=tslen/365), data=tg) +
  scale_color_viridis_c(begin=0.5) +
  theme_std() +
  labs(
    x="Longitude",
    y="Latitude",
    color="Timeseries Length (years)",
    size="Timeseries Length (years)",
    fill="EPA Level I Ecoregion"
  )
write.img(plt, "Figures/GageCovMap")
```

## Arkansas River Demo

This is the Arkansas River at Granite, CO, in between Leadville and Buena Vista
at about 9,000 ft.

```{r}
gn <- "07086000"
data <- read_csv("Data/GageData.csv",
                 col_types = "cdcdDid") %>%
  drop_na %>%
  filter(id == gn)
```

```{r}
data <- data %>%
  group_by(day) %>%
  mutate(tmean = mean(temperature)) %>%
  ungroup() %>%
  mutate(tsin = lm(tmean ~ cos((day - 210) * 2 * pi / 365))$fitted.values)
data <- left_join(data,
                  make.sints(fit.sins(data)),
                  by="day")
```
```{r}
plt <- ggplot(data, aes(day)) +
  geom_point(aes(y=temperature, color="Observations"),
             alpha=0.1, stroke=NA, size=2) +
  geom_line(aes(y=actemp, color="Three-Sine"), size=2) +
  geom_line(aes(y=tsin, color="Sine"), size=2) +
  scale_color_viridis_d(end=0.8) +
  theme_std() +
  labs(
    x="Day of Year",
    y="Gage Temperature (C)\n(Arkansas River at Granite, CO)",
    color=NULL
  )
write.img(plt, "Figures/ArkansasDemo", width=4)
```

## Conceptual Diagram

```{r}
# sysin <- tibble(
#   Intercept = 12,
#   Amplitude = 10,
#   FallDay = 330,
#   WinterDay = 40,
#   FallWinter = 3,
#   SpringDay = 150,
#   SummerDay = 230,
#   SpringSummer = 2
# )
sysin <- fit.sins(data)
ts <- make.sints(sysin) %>%
  rename(Value = actemp) %>%
  mutate(Component = "Overall Function (Sum)",
         Which = "All")
days <- ts$day
sinpt <-
  tibble(
    day = days,
    Value = sysin$Intercept + sysin$Amplitude * cos((day - 210) * 2 * pi / 365),
    Component = "Main Sine",
    Which = "Components"
  )
anoms <- mutate(ts,
               Value = Value - sinpt$Value,
               Which = "Components")
awin <- filter(anoms, day >= 250 | day <= 95) %>%
  mutate(Component = "Autumn/Winter Anomaly",
         Value = ifelse(abs(Value) == 0, NA, Value))
ssu <- filter(anoms, day > 95, day < 250, abs(Value) > 0) %>%
  mutate(Component = "Spring/Summer Anomaly")

inpt <- rbind(ts, sinpt, awin, ssu)
```

```{r}
plt <- ggplot(inpt, aes(day, Value, color=Component)) +
  geom_line(size=2) +
  facet_wrap(~Which, nrow=2) +
  scale_color_viridis_d(end=0.8) +
  theme_std() +
  labs(
    x="Day of Year",
    y="Mean Stream Temperature",
    color="Function Component"
  )
write.img(plt, "Figures/ConceptualDiagram")
```


## Coverage

```{r}
data <- left_join(data, tg, by=c("id"="site_no"))
```
```{r}
unq <- data %>%
  group_by(id) %>%
  slice_head(n=1) %>%
  ungroup()
```


```{r}
c(
  "East" = nrow(filter(unq, lon >= -100)) / nrow(unq),
  "Alaska" = nrow(filter(unq, lat > 50)),
  "Hawaii" = nrow(filter(unq, lon < -140, lat < 30))
)
```

```{r}
ecodata <- left_join(data, read_csv("Data/ecoregions.csv") %>%
                       select(id, ecoregion), by="id")
```
```{r}
ecodata %>%
  group_by(ecoregion) %>%
  summarize(
    N = length(unique(id))
  ) %>%
  ungroup() %>%
  mutate(Frac = N / sum(N))
```



