---
title: "Analysis of Stream Annual Temperature Cycles"
author: "Daniel Philippus"
date: "`r format(Sys.Date(), format='%b %d, %Y)')`"
output: pdf_document
---

# Setup

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE)
library(tidyverse)
source("functions.R")
source("util.R")
```

# Retrieve USGS Data

There are 67,706 gages with temperature data in US states.

## Identify Gages

```{r eval=FALSE}
all.tg <- map_dfr(state.abb, list.gages) %>%
  as_tibble %>%
  select(site_no,
         lat=dec_lat_va,
         lon=dec_long_va)
```
```{r eval=FALSE}
all.tg$tslen <- validate.gages(all.tg$site_no, log=T)
write_csv(all.tg, "Data/GageList.txt")
```

## Retrieve Data

### Gage List

886 gages have at least 10 years of data.  About 2900 have at least one year.
1500 have at least 5 years.  Maximum duration is 48 years.

```{r}
readgl <- TRUE

tg <- (if (readgl) {
  read_csv("Data/GageList.txt", col_types = "cddi")
  } else all.tg) %>%
  filter(tslen >= 365)

summary(tg)
```

#### Generate GEE Gage List

For retrieving data about gages.  Format: `[[lon, lat], "id"],`

```{r}
tg %>%
  mutate(
    r1 = paste0("[[", lon),
    r2 = paste0(lat, "]"),
    r3 = paste0('"', site_no, '"]')
  ) %>%
  select(r1:r3) %>%
  write_delim("EE/AllPts.txt", col_names = F, delim=",", eol=",\n", quote = "none", escape="none")
```



### Gage Properties

Of the ~1500 gages with 5 years' data, 1250 have non-NA information.

```{r}
gage.info <- (if (file.exists("Data/GageInfo.txt")) {
  read_csv("Data/GageInfo.txt", col_types = "cdcd")
} else {
  gage.info <- drop_na(get.gageinfo(tg$site_no))
  write_csv(gage.info, "Data/GageInfo.txt")
  gage.info
}) %>%
  group_by(id) %>%
  slice_head(n=1) %>%
  ungroup()

summary(gage.info)
```

### Temperature Records

Gages with available info and at least 5 years' data provide about 6 million rows,
with coverage from WY1965-present.  There are 1,234 gages with successfully 
retrieved data, making a mean of ~5000 days (14 years) per gage.

```{r eval=FALSE}
start <- "1950-01-01"
end <- "2023-01-01"

temps <- if (file.exists("Data/GageTemperatures.csv")) {
  read_csv("Data/GageTemperatures.csv", col_types = "cDid")
} else {
  gns <- gage.info$id
  is.done <- map_lgl(gns, ~file.exists(paste0("Data/GageChunks/", ., ".csv")))
  print(paste("Done:", sum(is.done)))
  cat("\n")
  for (gn in gns[!is.done]) {
    cat("\rRunning", gn)
    write_csv(get.usgs(gn, start, end), paste0("Data/GageChunks/", gn, ".csv"))
  }
  temps <- map_dfr(paste0("Data/GageChunks/", gns, ".csv"),
                   ~drop_na(read_csv(., col_types = "cDid")))
  write_csv(temps, "Data/GageTemperatures.csv")
  temps
}

summary(temps)
```

Quality control:

```{r}
tqc <- temps %>%
  filter(temperature >= 0,
         temperature <= 100)
```


## Full Dataset


```{r eval=FALSE}
data <- left_join(gage.info, tqc, by="id") %>%
  drop_na
write_csv(data, "Data/GageData.csv")
summary(data)
```

## Load Full Dataset

```{r}
data <- read_csv("Data/GageData.csv",
                 col_types = "cdcdDid") %>%
  drop_na
```

```{r}
data %>%
  group_by(id, day) %>%
  summarize(t = mean(temperature)) %>%
  summary
```

```{r}
data %>%
  group_by(id) %>%
  summarize(t = mean(temperature)) %>%
  summary
```

# Gage Timeseries Map

```{r}
plt <- plot.eco(TRUE) +
  scale_fill_viridis_d(option="magma", end=0.9) +
  geom_point(aes(lon, lat, size=tslen/365), data={
    left_join(data, tg, by=c("id"="site_no")) %>%
      group_by(id) %>%
      summarize(lon=first(lon),
                lat=first(lat),
                tslen=length(unique(Date)))
  }, color="yellow") +
  scale_color_viridis_c(begin=0.5) +
  theme_std() +
  labs(
    x="Longitude",
    y="Latitude",
    size="Timeseries Length (years)",
    fill="EPA Level I Ecoregion"
  )
write.img(plt, "Figures/GageCovMap_ts")
```

# Overview By Elevation

## Observed Rolling Means By Bucket

```{r}
plt <- bucketize(data, "elev", "m", 10) %>%
  group_by(id) %>%
  mutate(
    temp = temperature / mean(temperature)
  ) %>%
  group_by(BucketName, day) %>%
  summarize(
    temp = median(temp)
  ) %>%
  group_by(BucketName) %>%
  arrange(day) %>%
  mutate(temp = stats::filter(temp, rep(1/30, 30), circular=TRUE)) %>%
  ggplot() +
  aes(day, temp, color=BucketName) +
  geom_line(size=2) +
  scale_color_viridis_d(option="magma", end=0.8) +
  scale_x_continuous(breaks=30*(0:12)) +
  theme_std() +
  theme(legend.position = "bottom") +
  labs(x="Day of Year",
       y="Bucket Median Normalized Stream Temperature\n(30-Day Mean)",
       color="Elevation Bucket")

print(plt + theme_std(12))
write.img(plt, "Figures/ObservedElevationBuckets")
# ggsave("Figures/ObservedElevationBuckets.eps", plt, width=6.5, height=4)
```

## Model Data Preparation

```{r}
fitted <- data %>%
  sinu.resfit %>%
  drop_na() %>%
  bucketize("elev", "m", 10) %>%
  group_by(id) %>%
  group_modify(~{
    left_join(.x, make.sints(fit.sins(.x)), by="day")
    })
```

## Fourier Elevation Comparison

When `fixed` is TRUE (one extra variable), there is a very, very slight performance
penalty to the 3-sin model (<0.003 for all buckets and typ. <0.001 median).  When
`fixed` is FALSE (four extra variables), it has a small performance advantage for
all buckets except 185-229 m, but growing with elevation (up to 0.009).  Adding
a third harmonic (so two extra variables) reintroduces a small performance penalty
(0.002-0.003).

```{r warning=FALSE}
fitted <- data %>%
  group_by(id, day, elev) %>%
  summarize(temperature = mean(temperature)) %>%
  ungroup() %>%
  bucketize("elev", "m", 10) %>%
  group_by(id, BucketName) %>%
  arrange(day) %>%
  group_modify(~cbind(
    fit.sins(.x, fixed=FALSE),
    tibble(FR2 = fourier(.x$temperature, 3)$R2)))
```


```{r}
fitted %>%
  drop_na() %>%
  group_by(BucketName) %>%
  summarize(Gap = median(R2 - FR2),
            across(c(R2, FR2),
                   median))
```


## Residual Plot

```{r}
plt <- data %>%
  bucketize("elev", "m", 10) %>%
  group_by(BucketName, day) %>%
  summarize(
    temperature = mean(temperature)
  ) %>%
  group_by(BucketName) %>%
  arrange(day) %>%
  mutate(
    sinusoid = lm(temperature ~ cos((day - 210) * 2 * pi / 365))$fitted.values,
    residual = stats::filter(
      (temperature - sinusoid) / mean(temperature),
      rep(1/31, 31),
      circular=T
    )
  ) %>%
  ggplot(aes(day, residual, color=BucketName)) +
  geom_line(size=2) +
  scale_color_viridis_d(end=0.9) +
  theme_std() +
  theme(legend.position = "bottom") +
  labs(
    x="Day of Year",
    y="Mean Temperature Residual Compared to Sinusoid\n(31-Day Rolling Mean)",
    color="Elevation Bucket"
  )
write.img(plt, "Figures/Residual")
```



## Comparison Plot

This shows the point very well.  Model behavior is nearly indistinguishable at
low elevation, while the anomaly model clearly fits observed behavior better at
high elevation.  However, it is evident that even at lower elevations the anomaly
model does better; it is just a subtle enough effect not to be obvious unless
specifically looked for.

```{r}
plt <- fitted %>%
  group_by(BucketName, day) %>%
  summarize(
    across(c(ntemp, reftemp, RefMod, actemp),
           ~median(.))
  ) %>%
  group_by(BucketName) %>%
  mutate(actemp = actemp / mean(actemp)) %>%
  arrange(day) %>%
  mutate(
    across(c(ntemp, reftemp, RefMod, actemp),
           ~stats::filter(., rep(1/30, 30), circular=TRUE))
  ) %>%
  rename(Sinusoidal = reftemp,
         `Anomaly Model` = RefMod,
         `Three-Sinusoid Model` = actemp) %>%
  pivot_longer(c(Sinusoidal, `Anomaly Model`, `Three-Sinusoid Model`),
               names_to="Model",
               values_to="Prediction") %>%
  ggplot() +
  aes(day) +
  geom_line(aes(y=Prediction, color=Model), size=1) +
  geom_point(aes(y=ntemp, color="Observed"), size=1) +
  facet_wrap(~BucketName, nrow=2) +
  scale_color_viridis_d() +
  theme_std() +
  labs(
    x="Day of Year",
       y="Bucket Median Normalized Stream Temperature\n(30-Day Mean)",
       color="Model Type"
  )

print(plt + theme_std(12))
write.img(plt, "Figures/ElevationModelComparison")
```

## Model Performance Comparison

```{r}
perf <- fitted %>%
  drop_na() %>%
  group_by(id, day) %>%
  summarize(
    across(c(ntemp, reftemp, ac.3sin, ac.5sin),
           mean),
    BucketName = first(BucketName)
  ) %>%
  group_by(id) %>%
  summarize(
    BucketName = first(BucketName),
    across(c(reftemp, ac.3sin, ac.5sin),
           ~cor(.x, ntemp)^2),
    Delta3 = ac.3sin - reftemp,
    Delta5 = ac.5sin - reftemp
  ) %>%
  ungroup()
```

The anomaly model outperforms the sinusoid model for about 3/4 of gages.  The
typical difference is modest, but it closes about half the small gap to perfect
performance.  Importantly, the anomaly model does much better for some marginally
worse-performing areas (e.g. R2 around 0.9-0.95).

The performance gap grows quite rapidly with elevation buckets.  From 0-60%,
the gap in R2 is under 0.02, growing by around 0.001-0.003 per bucket.  It then
grows quite rapidly up to 90%, maxing out at a gap of almost 0.05.  This mirrors
the visible behavior in the plots.

```{r}
summary(select(perf, reftemp, ac.3sin, Delta3))
```
```{r}
perf %>%
  group_by(BucketName) %>%
  summarize(
    across(c(reftemp, starts_with("ac"), starts_with("Delta")),
           median)
  )
```

## Statistically Significant Difference

Test differences between the mean and the sinusoid (`reftemp`) and 3sin (`actemp`)
temperatures.

```{r}
difs <- fitted %>%
  group_by(id) %>%
  mutate(sinusoid = reftemp * mean(temperature)) %>%
  group_by(id, day) %>%
  filter(n() >= 30) %>%
  summarize(
    Mean = mean(temperature),
    SD = sd(temperature),
    across(c(sinusoid, actemp),
           list(
             "zscore" = ~(mean(.x) - Mean) / (SD / sqrt(n())),
             "pval" = ~pnorm(mean(.x), mean=Mean, sd=SD/sqrt(n()), lower.tail=(.x <= Mean))
           ))
  )
```
```{r}
summary(select(ungroup(difs), Mean:actemp_pval))
```
```{r}
difs %>%
  group_by(id) %>%
  summarize(
    across(
      ends_with("_pval"),
      ~sum(.x < 0.05) / n()
    )
  ) %>%
  summary
```



# Three-Sinusoid Fit Testing

Use high elevation to test.

```{r}
uel <- data %>% bucketize("elev", "m", 10) %>% filter(Bucket == 90)
```

We have the usual visibly non-sinusoidal shape.

```{r}
uel.mean <- uel %>%
  group_by(day) %>%
  summarize(
    temperature = mean(temperature, na.rm=T)
  )
plot(uel.mean$day, uel.mean$temperature)
```

```{r}
sinft <- fit.sins(uel.mean)
fitted <- uel.mean %>%
  add.sinusoid.all %>%
  add.resfit %>%
  left_join(make.sints(sinft),
            by="day")
```

The sinusoidal model has the usual shortcomings.

```{r}
plot(fitted$day, fitted$temperature)
lines(fitted$day, fitted$reftemp * mean(fitted$temperature), col="red")
```
The anomaly model, on the other hand, does quite well, as per usual.

```{r}
plot(fitted$day, fitted$temperature)
lines(fitted$day, fitted$RefMod * mean(fitted$temperature), col="red")
```

The three-sinusoid model also does very well and behaves quite similarly.
They trade off doing well or poorly, but the 3S model does better at the maximum
and minimum.

```{r}
plot(fitted$day, fitted$temperature)
lines(fitted$day, fitted$actemp, col="red")
lines(fitted$day, fitted$RefMod * mean(fitted$temperature), col="green")
```

The 3S model is very marginally the best performer, and both solidly outdo the
sinusoid model.

```{r}
cor(fitted$temperature, fitted$reftemp)^2
cor(fitted$temperature, fitted$RefMod)^2
cor(fitted$temperature, fitted$actemp)^2
```

# Demo Plots

Replicate preliminary figures from TE2 seasonality work: modeled anomalies.

```{r}
fits <- data %>%
  bucketize("elev", "m", 20) %>%
  group_by(BucketName, day) %>%
  summarize(temperature = mean(temperature)) %>%
  group_by(BucketName) %>%
  group_modify(~{
    fit <- fit.sins(.x)
    prd <- make.sints(fit)
    tibble(
      day = prd$day,
      actemp = prd$actemp - (fit$Intercept + fit$Amplitude * cos((day - 210) * 2 * pi / 365))
    )
  })
```

```{r}
plt <- ggplot(fits, aes(day, actemp, color=BucketName)) +
  geom_line(size=2) +
  scale_color_viridis_d() +
  theme_std() +
  theme(legend.position = "bottom") +
  labs(
    x="Day of Year",
    y="Predicted Seasonal Temperature Residual",
    color="Elevation\nBucket"
  )
write.img(plt, "Figures/ElevationAnomFitted")
```




# Formally Characterizing Model Differences

## lm Statistical Significance

```{r}
lms <- data %>%
  group_by(id) %>%
  group_modify(~{
    safely(\(x) summary(fit.sins(x, TRUE))$coefficients[,4] %>%
      as.list() %>%
      as_tibble,
      otherwise=tibble(sin1=NA, sin2=NA))(.)$result
  })
```
```{r}
summary(select(lms, -id))
```

```{r}
quantile(lms$sin1, 1:100 / 100, na.rm=T)
```
```{r}
quantile(lms$sin2, 1:100 / 100, na.rm=T)
```

```{r}
lms <- data %>%
  bucketize("elev", "m", 10) %>%
  group_by(BucketName) %>%
  group_modify(~{
    safely(\(x) summary(fit.sins(x, TRUE))$coefficients[,4] %>%
      as.list() %>%
      as_tibble,
      otherwise=tibble(sin1=NA, sin2=NA))(.)$result
  })
```
```{r}
lms
```

## Compare AIC

Use `fit.sins(mod=TRUE, aic=TRUE)` to compute AIC.

```{r}
aicda <- data %>%
  group_by(id) %>%
  group_modify(~fit.sins(.x, TRUE, TRUE)) %>%
  mutate(Delta = Anom - Sin) %>%
  ungroup() %>%
  select(-id)
```
```{r}
summary(aicda)
quantile(aicda$Delta, (1:100)/100)
```
```{r}
data %>%
  bucketize("elev", "m", 10) %>%
  group_by(BucketName) %>%
  group_modify(~fit.sins(.x, TRUE, TRUE)) %>%
  mutate(Delta = Anom - Sin) %>%
  mutate(
    across(Sin:Delta, as.integer)
  ) %>%
  rename(
    `Elevation Bucket`=BucketName,
    `Sinusoidal Model AIC`=Sin,
    `Three-Sine Model AIC`=Anom,
    `AIC Difference (Three-Sine minus Sinusoidal)`=Delta
  ) %>%
  write_delim("AICtab.txt",
              delim="&",
              eol="\\\\\n",
              quote="none")
```

## Compare to Fourier Transform

### 3S Model Performance Distribution

```{r}
fitted <- data %>%
  group_by(id) %>%
  filter(length(unique(day)) >= 365) %>%
  group_modify(~fit.sins(.x, fixed=TRUE)) %>%
  ungroup()
```
```{r}
summary(fitted$R2)
summary(fitted$RMSE)
```
### Fourier Transform

```{r message=FALSE}
foured <- data %>%
  group_by(id, day) %>%
  summarize(temperature = mean(temperature)) %>%
  group_by(id) %>%
  filter(length(unique(day)) >= 365) %>%
  summarize(fourier(temperature, 2))

summary(foured)
```

Note that the same amount of data is required when N=4 (real and imaginary part).

For a median R2 of 0.98-0.99 (IQR 0.97-0.99) and RMSE of 0.78 C (0.58-1.07), about
two terms are required (four numbers).  That's in comparison to fitting all four
dates (eight numbers).  With `fixed=TRUE` (five numbers), it is 0.98 (0.96-0.99)
and 0.87 (0.65-1.19), which still requires N=2 (five numbers vs four).

An alternative approach is asymmetrical, since the first Fourier term doesn't
wrap around.  However, this doesn't affect the behavior much.


# Using the Model - For Science!

```{r}
data <- read_csv("Data/GageData.csv",
                 col_types = "cdcdDid") %>%
  drop_na
```

## Geographic Clusters (Riverwise)

```{r}
rivers <- data %>%
  group_by(id) %>%
  group_modify(~cbind(.x, fit.sins(.x))) %>%
  summarize(
    across(da:elev, first),
    across(c(temperature, Intercept:R2),
           mean)
  ) %>%
  left_join(tg, by=c("id"="site_no"))
```

This is a great set of plots for showcasing the added information from the other
seasonality parameters.  Many have clear spatial signals which do not align with
the temperature-amplitude model.

- Mean temperature is high in the south, at low elevation, and near the Gulf
(duh).
- Amplitude is high inland and north (duh).

And then the new information:

- The fall low is strong in the inland north, like amplitude, but it doesn't
weaken as much as amplitude in the southern coastal regions.  Its strong areas
also show a firmer boundary rather than a more gradual shift.
- The timing of the spring low has a quite different pattern.  It hits earliest
in the northeastern quarter of the country, latest in the southeast, and towards
the middle in the west, though it is quite variable throughout.
- The magnitude of the spring low, meanwhile, seems to be driven by elevation.  It
is strongest in the Rockies and weakest in the low-elevation Southeast.  This
corresponds to the observed seasonality plot.
- Summer highs are strongest (but variable) in the west and weaker in the east.
It seems likely this corresponds to the wet-winter/dry-summer regime typical
of the west.
- The winter high hits earliest further north and on the coasts, while it hits
later to the south and inland.  This must have something to do with cold ocean influence,
though it is unclear why this would affect timing.
- The magnitude of the winter high is rather variable geographically, but it tends
to be strongest inland and weaker on the coasts.  Particularly strong points seem
to be concentrated in the western Rockies and eastern Sierra or in the parts of the
Basin and Range close to those mountain ranges.


```{r}
states <- map_data("state")
plt <- rivers %>%
  group_by(id) %>%
  group_modify(~normalize.sins(.x)) %>%
  ungroup() %>%
  filter(lon > -130) %>%
  pivot_longer(c(temperature, Amplitude:SpringSummer),
               names_to="Aspect",
               values_to="Value") %>%
  group_by(Aspect) %>%
  mutate(Value = (Value - mean(Value))/sd(Value)) %>%
  ggplot() +
  geom_polygon(aes(long, lat, group=group), data=states, color="grey",
               fill=NA, size=2) +
  geom_point(aes(lon, lat, color=Value), size=2) +
  scale_color_viridis_c(limits=c(-3, 3)) +
  facet_wrap(~Aspect, nrow=2, labeller=as_labeller(renamer)) +
  theme_std() +
  labs(
    x="Longitude",
    y="Latitude",
    color="Normalized\nSeasonality\nParameter"
  )
print(plt + theme_std(12))
write.img(plt, "Figures/RiverSeasonMaps")
```

A final point to investigate would be clustering.

### Formal Clustering

Using k-means and the following cluster counts: (note that there are always
numerous exceptions)

2. Cluster 2 dominates the west coast mountains, Northeast, and Great Lakes, with
a significant minority presence in the Rockies.  Cluster 1 dominates elsewhere.
3. Cluster 2 is divided into Clusters 1 and 2, with 2 taking over in the Cascades.
Cluster 3 overlaps well with the former Cluster 1.
4. Cluster 3 now divides into Clusters 3 and 4.  3 is more dominant in the west.
5. The previous pattern is largely unchanged.  The new Cluster 5 is widely but
sparsely distributed.

6 onwards provide little new information.  From 2-4, it appears that key regions are:

1. West Coast mountains, Northeast, and Great Lakes; from which the Cascades,
especially the northern Cascades, can be separated.
2. South and inland North, which can be further divided into the Rockies plus the
Southeast versus the Midwest and the rest of the East.

It may be interesting to generate plots of the mode in a moving window for the 
k=2 and k=4 maps (or just k=4).

```{r}
for (k in 2:10) {
  rivers$cluster <- as.factor(kmeans(select(rivers, temperature, Amplitude:SpringSummer),
                           k)$cluster)
  plt <- ggplot(filter(rivers, lon > -130), aes(lon, lat, color=cluster)) +
    geom_polygon(aes(long, lat, group=group), data=states, color="grey",
               fill=NA, size=2) +
    geom_point(size=2) +
    scale_color_viridis_d() +
    theme_std() +
    labs(
      x="Longitude",
      y="Latitude",
      color="Seasonality\nCluster"
    )
  write.img(plt, paste0("Figures/RiverwiseClusters/Clusters_k", k))
}
```

Unfortunately, this summary approach is quite noisy.  Best to stick to individual
points.


```{r}
rivers$cluster <- as.factor(kmeans(select(rivers, temperature, amplitude, WX:FY),
                           4)$cluster)
```

```{r}
getmode <- function(v) {
   uniqv <- unique(v)
   uniqv[which.max(tabulate(match(v, uniqv)))]
}

plt <- ggplot() +
  geom_polygon(aes(long, lat, group=group), data=states, color="black",
               fill=NA, size=2) +
  stat_summary_2d(aes(lon, lat, z=cluster), data=filter(rivers, lon > -130),
                  fun=getmode, bins=20, alpha=0.8) +
  scale_fill_viridis_d() +
  theme_std() +
  labs(
    x="Longitude",
    y="Latitude",
    fill="Dominant\nSeasonality\nCluster"
  )
write.img(plt, "Figures/RiverwiseClusters/ModalCluster")
```

### Analysis of River Trends

```{r}
rivers %>%
  pivot_longer(
    c(Intercept, Amplitude, FallWinter, SpringSummer),
    names_to="Ssn",
    values_to="Sv"
  ) %>%
  pivot_longer(c(lat, lon, elev, da),
               names_to="Riv",
               values_to="Rv") %>%
  ggplot(aes(Rv, Sv)) +
  geom_point(alpha=0.1, stroke=NA) +
  theme_bw() +
  facet_grid(Ssn ~ Riv,
             switch="both",
             scales="free") +
  labs(x=NULL, y=NULL)
```



## Geographic Trends

### Moving Window

Compute overall seasonality for all rivers within 2 degrees centered on each
integer coordinate pair.  Plot each trend.

```{r}
geo <- filter(left_join(data, tg, by=c("id"="site_no")), lon >= -130) %>%
  drop_na
```

```{r}
lons <- (-130):(-65)
lats <- 25:50
width <- 2

geossn <- map_dfr(lons,
                  function(x) {
                    map_dfr(lats,
                            function(y) {
                              raw <- filter(geo,
                                            lon >= x - width/2,
                                            lon < x + width/2,
                                            lat <= y + width/2,
                                            lat > y - width/2)
                              if (nrow(raw) > 0) {
                                cbind(
                                  tibble(
                                    lon = x,
                                    lat = y,
                                    temperature = mean(raw$temperature)
                                  ),
                                  fit.sins(raw)
                                ) %>%
                                  normalize.sins()
                              } else tibble()
                            })
                  })
```


Some variables range quite substantially by location.  For example, the spring
low and winter high can range from negligible to a large fraction of an amplitude.

```{r}
summary(geossn)
```

In these maps:

- Temperatures are high south, low, and near the Gulf.
- Amplitudes are high north, inland, and in Maine.
- The autumn/winter anomaly is large at high elevation and in the northern Great Plains.
The autumn low hits early where amplitude is high, and the winter anomaly hits early
near cold water.  The anomaly can reach up to 20-40% of mean temperatures, which is
a much larger proportion of mean temperatures in the fall/winter.
- The spring/summer anomaly is strongest in the high-elevation west and Maine.
 The spring low hits early around the Great Lakes, and the summer high hits early
 inland.  The anomaly can likewise reach 20-40% of mean temperatures.

```{r}
geossn <- select(geossn, -any_of(c("R2", "Intercept")))
states <- map_data("state")
for (nm in names(geossn)[-(1:2)]) {
  plt <- ggplot(geossn,
                aes(lon, lat, fill=.data[[nm]])) +
    geom_polygon(aes(long, lat, group=group), data=states, color="black",
                 fill=NA, size=2) +
    geom_tile(alpha=0.9) +
    scale_fill_viridis_c() +
    theme_std() +
    labs(
      x="Longitude",
      y="Latitude",
      fill=paste0("Window\n", renamer[nm])
    )
  write.img(plt, paste0("Figures/MovingWindowSeasons/", nm))
}

plt <- geossn %>%
  mutate(
    across(temperature:SpringSummer,
           ~(.x - mean(.x))/sd(.x))
  ) %>%
  pivot_longer(temperature:SpringSummer,
               names_to="Aspect",
               values_to="Value") %>%
  ggplot(aes(lon, lat, fill=Value)) +
  geom_polygon(aes(long, lat, group=group), data=states, color="black",
                 fill=NA, size=2) +
  geom_tile(alpha=0.9) +
  scale_fill_viridis_c(limits=c(-3, 3)) +
  facet_wrap(~Aspect, nrow=2, labeller = as_labeller(renamer)) +
  theme_std() +
  labs(
    x="Longitude",
    y="Latitude",
    fill="Normalized\nSeasonality\nTerm"
  )
write.img(plt, "Figures/MovingWindowSeasons/All")
```

### Elevation (100 m)

```{r}
width <- 100

elev <- data %>%
  group_by(elev = width * floor(elev / width), id) %>%
  group_modify(~normalize.sins(fit.sins(.))) %>%
  group_by(elev) %>%
  summarize(
    across(-id,
           list(
             "median" = median,
             "outliers" = ~sum(abs((.x - mean(.x)) / sd(.x)) >= 3)
           )),
    outliers = Intercept_outliers | Amplitude_outliers | FallWinter_outliers | SpringSummer_outliers,
    N = n()
  ) %>%
  rename_with(\(x) str_split(x, "_", simplify=T)[,1], ends_with("_median"))
  # group_modify(~mutate(
  #   normalize.sins(fit.sins(.)), N = length(unique(.$id))
  # ))
```

```{r}
ol.elevs <- c(
                2100,
                2500,
                2600,
                2800)
```

```{r}
olel.names <- c(
  "2100" = "Yellowstone and Eastern Sierra",
  "2500" = "Northern Colorado Rockies",
  "2600" = "Northern/Western Colorado Rockies and Northwestern Wyoming",
  "2800" = "Western/Southwestern Colorado Rockies"
)
```


```{r}
olel <- 2800

left_join(filter(data, elev >= olel, elev < olel + 100),
          tg, by=c("id"="site_no")) %>%
  ggplot() +
  geom_polygon(aes(long, lat, group=group), data=states, fill=NA, color="black") +
  geom_point(aes(lon, lat))
```


```{r}
outliers <- elev %>%
  pivot_longer(c(Intercept, Amplitude, FallWinter, SpringSummer)) %>%
  filter(
    elev %in% ol.elevs
  ) %>%
  mutate(Which = olel.names[as.character(elev)])
```


```{r}
plt <- elev %>%
  pivot_longer(c(Intercept, Amplitude, FallWinter, SpringSummer)) %>%
  ggplot(aes(elev, value, shape=N < 10, group=1)) +
  geom_point(size=4) +
  geom_point(aes(elev, value, shape=N < 10, color=Which), data=outliers, size=8) +
  scale_color_viridis_d(end=0.9, guide=guide_legend(nrow=3,
                                                    title.position = "top",
                                                    override.aes = list(shape=17))) +
  geom_smooth(size=2, method="lm", se=T) +
  facet_wrap(~name, scales="free_y", labeller = as_labeller(renamer),
             strip.position = "left", nrow=2) +
  theme_std() +
  theme(legend.position = "bottom") +
  labs(
    x="Elevation (m)",
    y=NULL,
    shape="Few Rivers (<10)",
    color="Outlier Contents"
  )
write.img(plt, "Figures/ElevationTrends")
```




## Temporal Trends (Rolling Mean)

### Long Term Data

```{r}
# minyr <- 2010

datats <- map_dfr(c(1990, 2000, 2010),
                  function(minyr) {
                    data %>%
                      drop_na() %>%
                      group_by(id) %>%
                      mutate(Year = as.integer(format(Date, "%Y")),
                             MinYear = min(Year),
                             MaxYear = max(Year)) %>%
                      ungroup() %>%
                      filter(MinYear <= minyr,
                             MaxYear >= 2022) %>%
                      filter(Year >= minyr,
                             Year <= 2022) %>%
                      mutate(StartYr = minyr)
                  })

datats %>%
  group_by(StartYr) %>%
  summarize(N = length(unique(id)))
```


### By River

```{r warning=FALSE}
window <- 10  # years
trends <- datats %>%
  group_by(id, day) %>%
  arrange(Date) %>%
  mutate(
    temperature = zoo::rollmean(temperature, window, fill=NA, align="left")
  ) %>%
  ungroup() %>%
  drop_na() %>%
  group_by(id, Year = as.integer(format(Date, "%Y"))) %>%
  arrange(Date) %>%
  group_modify(~{
    tryCatch({add.resfit(add.sinusoid(.x)) %>%
      summarize(
        across(da:elev, first),
        temperature = mean(temperature),
        across(c(amplitude, WX:FY),
               mean),
        R2 = cor(RefMod, cycle.temp)^2
      )
    },
    error=function(e) tibble())}, .keep=TRUE) %>%
  drop_na()
```

```{r}
tcor.wide <- trends %>%
  group_by(id) %>%
  summarize(
    across(temperature:FY,
           ~safely(cor, otherwise=0)(Year, ., method="spearman")$result)
  ) %>% drop_na
```


```{r}
tcors <- trends %>%
  pivot_longer(temperature:FY,
               names_to="Aspect",
               values_to="Value") %>%
  group_by(id, Aspect) %>%
  summarize(
    Cor = safely(cor, otherwise=0)(Year, Value, method="spearman")$result,
    p = safely(cor.test, otherwise=list(p.value=1))(Year, Value, method="spearman")$result$p.value,
    len = n()
  ) %>%
  filter(p < 0.05)

tcors %>%
  arrange(-abs(Cor))
```

Compare what we see with just mean-amplitude versus what we see with more data.
Also make a map.

#### Histogram

```{r}
plt <- ggplot(tcors, aes(Cor)) +
  geom_histogram(bins=10) +
  facet_wrap(~Aspect, labeller = as_labeller(\(x) paste0("Cor(Start Year, ", renamer[x], ")")),
             strip.position = "bottom", nrow=2) +
  theme_std() +
  labs(
    x=NULL,
    y="River Count"
  )

print(plt + theme_std(12))
write.img(plt, "Figures/SignificantCorrelations_Temporal_ByRiver")
```

#### Cross-Scatterplots

To illustrate, make scatterplots of correlations against each other.  It is very
interesting to see. Some have no visible correlation between correlations; for
example, most rivers are increasing in mean temperature, but many of those do
*not* have a further increasing summer high.  On the other hand, quite strong
cross-correlations do exist.  Strengthening spring lows correlate with strengthening
winter highs, which in turn correlate with strengthening fall lows.  Strengthening
fall lows also correlate with later spring lows.  And so on.

Note that lows (fall/spring) are strengthening with a negative correlation.

```{r}
nms <- names(tcor.wide)[-1]
for (nm1 in nms) {
  for (nm2 in nms[nms > nm1]) {
    x <- tcor.wide[[nm1]]
    y <- tcor.wide[[nm2]]
    plt <- ggplot(mapping = aes(x, y)) +
      geom_point(size=2) +
      theme_std() +
      labs(x=paste0("Cor(Start Year, ", renamer[nm1], ")"),
           y=paste0("Cor(Start Year, ", renamer[nm2], ")"))
    write.img(plt, paste0("Figures/RiverTemporalCorScatters/", nm1, "_", nm2))
  }
}
```


Of the 28 combinations, 22 have statistically significant cross-correlations.

Of note:

- (SuY, amplitude) - temperature: negative - warming rivers have a weaker summer high
and weaker swings in general.
- (amplitude, SuY) - WY: positive - more variable rivers with a warmer summer have a
stronger winter high.
- FY - (temperature, WY): (negative, positive) - rivers with a deeper fall low
are warmer but have a weaker winter high
- SpY - WY: positive - rivers with a stronger spring low have a weaker winter high
- FY - SpY: positive - both strengthen

```{r}
nms <- names(tcor.wide)[-1]
xcor <- map_dfr(nms,
                function(nm1) {
                  map_dfr(nms[nms > nm1],
                          function(nm2) {
                            tibble(Var1=nm1,
                                   Var2=nm2,
                                   XCor=cor(tcor.wide[[nm1]],
                                            tcor.wide[[nm2]],
                                            method="spearman"),
                                   p=cor.test(tcor.wide[[nm1]],
                                            tcor.wide[[nm2]],
                                            method="spearman")$p.value)
                          })
                })
arrange(xcor, -abs(XCor))
```


#### Map

Must load `tg` from data retrieval for coordinates.

```{r}
tc.info <- trends %>%
  group_by(id) %>%
  summarize(
    across(c(da:elev), first),
    across(temperature:FY,
           ~safely(cor, otherwise=0)(Year, ., method="spearman")$result)
  ) %>%
  drop_na %>%
  left_join(
    select(tg, -tslen),
    by=c("id"="site_no")
  )
```

Assess general geographic trends.  There are notable latitude correlations with
WX trends, longitude with FY, temperature, amplitude, and SpY trends, and altitude
with WX trends.

Longitude correlates with most things, so it's worth plotting all of them.

```{r}
tc.info %>%
  summarize(across(temperature:FY,
                   ~cor(., lon)))
```

In general, regional trends in time overlap relatively well between the sinusoid
model and the anomaly model.  It is at a more local scale where differences become
evident.

```{r}
states <- map_data("state")

plt <- tc.info %>%
  filter(lat < 50) %>%
  pivot_longer(temperature:FY,
               names_to="Aspect",
               values_to="Cor") %>%
  ggplot() +
  geom_polygon(aes(long, lat, group=group), data=states, size=2, color="grey",
               fill=NA) +
  stat_summary_2d(aes(lon, lat, z=Cor), bins=20) +
  scale_fill_viridis_c() +
  facet_wrap(~Aspect, nrow=2, labeller=as_labeller(renamer)) +
  theme_std() +
  labs(
    x="Longitude",
    y="Latitude",
    fill="Cor(Start Year, Variable)"
  )

print(plt + theme_std(12))
write.img(plt, "Figures/RiverTemporalCorMap")
```

#### Clustering

2. No obvious spatial trend.
3. Cluster 2 is relatively rare inland and more common on the west coast.
4. Similarly (now Cluster 4).
Etc.

It appears that there is one sort of trend that is prevalent nearer the coasts,
and otherwise no obvious pattern.

```{r}
states <- map_data("state")
for (k in 2:10) {
  tc.info$cluster <- as.factor(kmeans(select(tc.info, temperature:FY),
                           k)$cluster)
  plt <- ggplot(filter(tc.info, lon > -130), aes(lon, lat, color=cluster)) +
    geom_polygon(aes(long, lat, group=group), data=states, color="grey",
               fill=NA, size=2) +
    geom_point(size=2) +
    scale_color_viridis_d() +
    theme_std() +
    labs(
      x="Longitude",
      y="Latitude",
      color="Seasonality\nTemporal Trend\nCluster"
    )
  write.img(plt, paste0("Figures/RiverTrendClusters/Clusters_k", k))
}
```


### By Elevation Bucket

This reveals distinctive trends.  For example, at high elevation:

- Amplitude has been roughly constant.  Lower elevations saw a dip around 1995
and then recovery.
- Temperature has also been roughly constant.  It has increased for the bottom ~60%.
- The fall low has weakened, but only for the highest elevations.
- The spring low has gotten stronger and a little earlier for the highest elevations.
- The summer high has grown slightly, while at lower elevations it has generally
shrunk a bit.
- The winter high has weakened slightly for most elevations.

```{r warning=FALSE}
window <- 5  # years
trends <- datats %>%
  bucketize("elev", "m", 10) %>%
  group_by(StartYr) %>%
  group_modify(~{
    .x %>%
      group_by(BucketName, Date, day) %>%
      summarize(
        temperature = median(temperature)
      ) %>%
      group_by(BucketName, day) %>%
      arrange(Date) %>%
      mutate(
        temperature = zoo::rollmean(temperature, window, fill=NA, align="right")
      ) %>%
      ungroup() %>%
      drop_na() %>%
      group_by(BucketName, Year = as.integer(format(Date, "%Y"))) %>%
      arrange(Date) %>%
      group_modify(\(z, w) safely(\(x) normalize.sins(
        mutate(
          fit.sins(x),
          temperature = mean(x$temperature)
        )
      ),
                           otherwise=tibble())(z)$result, .keep=TRUE) %>%
      drop_na()
  })
```

```{r}
plt <- trends %>%
  pivot_longer(c(Intercept, Amplitude, FallWinter, SpringSummer),
               names_to="Aspect",
               values_to="Value") %>%
  ggplot(aes(Year, Value, color=as.factor(StartYr))) +
  geom_line(size=2) +
  # geom_smooth(size=2, se=F) +
  scale_color_viridis_d() +
  facet_grid(Aspect ~ BucketName, switch="y", scales="free_y",
             labeller=labeller(Aspect=as_labeller(renamer))) +
  theme_std() +
  scale_x_continuous(breaks=c(2000, 2020)) +
  theme(legend.position = "bottom") +
  labs(
    x="End Year",
    y=NULL,
    color="Start Year"
  )
# print(plt + theme_std(12))
write.img(plt, "Figures/ElevationBucketTrends")
```


### US

For the USA as a whole:

- Temperature has been gradually increasing.
- Amplitude has been increasing after a sharp drop in the 1990s.
- The fall low has been getting a bit stronger.
- The spring low has been variable and small.
- The summer high has been variable and small.
- The winter high has been weakening consistently.

```{r}
window <- 5  # years
trends <- datats %>%
  group_by(StartYr, Date, day) %>%
  summarize(temperature = median(temperature)) %>%
  group_by(StartYr, day) %>%
  arrange(Date) %>%
  mutate(
    temperature = zoo::rollmean(temperature, window, fill=NA, align="right")
  ) %>%
  ungroup() %>%
  drop_na() %>%
  group_by(StartYr, Year = as.integer(format(Date, "%Y"))) %>%
  arrange(Date) %>%
  group_modify(~safely(fit.sins, otherwise=tibble())(.)$result, .keep=TRUE)
```

```{r}
plt <- trends %>%
  pivot_longer(c(Intercept, Amplitude, SpringSummer, FallWinter),
               names_to="Aspect",
               values_to="Value") %>%
  ggplot(aes(Year, Value, color=as.factor(StartYr))) +
  scale_color_viridis_d(end=0.9) +
  geom_line(size=4) +
  # geom_smooth(size=2, color="red", linetype="dashed", se=F) +
  facet_wrap(~Aspect, nrow=1, strip.position = "left", scales="free_y",
             labeller=as_labeller(renamer.nn)) +
  theme_std() +
  theme(legend.position = "bottom") +
  labs(
    x="End Year",
    y=NULL,
    color="Start Year"
  )
print(plt + theme_std(12))
write.img(plt, "Figures/USATrendsx3", height=2)
```

#### US by Grid Cell

Note: add *n* to legend and convert plots to be anomaly relative to 1994.

```{r}
plotpts <- datats %>%
  filter(StartYr == 1990, Year == 1994) %>%
  left_join(tg, by=c("id"="site_no")) %>%
  group_by(id) %>%
  slice_head(n=1)

plot.eco() +
  geom_point(aes(lon, lat), data=plotpts, color="yellow")
```


```{r}
window <- 5  # years
trends <- datats %>%
  filter(StartYr == 1990) %>%
  left_join(tg, by=c("id"="site_no")) %>%
  group_by(LonGrp = 10 * floor(lon/10),
           # LatGrp = 10 * floor(lat/10)
           ) %>%
  mutate(N = length(unique(id))) %>%
  # filter(LonGrp %in% c(-100, -130)) %>%
  group_by(StartYr, Date, day,
           # LatGrp = 10 * floor(lat/10),
           LonGrp,
           # LatGrp,
           N) %>%
  summarize(temperature = median(temperature, na.rm=T)) %>%
  group_by(StartYr, day,
           # LatGrp,
           LonGrp,
           # LatGrp
           ) %>%
  arrange(Date) %>%
  mutate(
    temperature = zoo::rollmean(temperature, window, fill=NA, align="right")
  ) %>%
  ungroup() %>%
  drop_na() %>%
  group_by(StartYr, Year = as.integer(format(Date, "%Y")),
           # LatGrp,
           LonGrp,
           # LatGrp,
           N
           ) %>%
  arrange(Date) %>%
  group_modify(~safely(fit.sins, otherwise=tibble())(.)$result, .keep=TRUE) %>%
  group_by(LonGrp
           # ,LatGrp
           ) %>%
  mutate(Has1994 = 1994 %in% Year) %>%
  ungroup() %>%
  filter(Has1994)
```

```{r}
plt <- trends %>%
  pivot_longer(c(Intercept, Amplitude, SpringSummer, FallWinter),
               names_to="Aspect",
               values_to="Value") %>%
  group_by(Aspect, LonGrp
           # ,LatGrp
           ) %>%
  arrange(Year) %>%
  mutate(Value = Value - first(Value)) %>%
  ggplot(aes(Year, Value, color=as.factor(-LonGrp),
             size=N)) +
  scale_color_viridis_d(end=0.9) +
  scale_size_continuous(range=c(2,8), breaks=c(1, 1:5*10)) +
  geom_line(alpha=0.8) +
  # geom_smooth(size=2, color="red", linetype="dashed", se=F) +
  facet_wrap(~Aspect, nrow=2, strip.position = "left", scales="free_y",
             labeller=as_labeller(renamer.nn)) +
  theme_std() +
  guides(color=guide_legend(override.aes = list(size=12))) +
  theme(legend.position = "bottom") +
  labs(
    x="End Year",
    y="Anomaly Relative to 1994",
    color="Longitude Group (West Edge; deg W)",
    size="Gage Count"
  )
print(plt + theme_std(12))
write.img(plt, "Figures/USATrendsLon")
```

#### Texas Test

```{r}
datats %>%
  left_join(tg, by=c("id"="site_no")) %>%
  filter(lon > -105,
         lon < -95,
         lat < 35) %>%
  group_by(StartYr,
           Year) %>%
  summarize(temperature = median(temperature)) %>%
  ggplot(aes(Year, temperature, color=as.factor(StartYr))) +
  theme_bw() +
  geom_line()
```


# Case Study Validation

## Temporal Trends

### USA: 2000 dataset, 2010-2020, autumn (315) down/winter (40) up 0.4 C

```{r}
datats %>%
  group_by(StartYr, Year, day) %>%
  summarize(temperature = mean(temperature)) %>%
  group_by(StartYr, day) %>%
  arrange(Year) %>%
  mutate(temperature = zoo::rollmean(temperature, 5, fill=NA, align="right")) %>%
  ungroup() %>%
  filter(StartYr == 2000, Year %in% c(2010, 2020), day %in% c(40, 315)) %>%
  group_by(Year, day) %>%
  summarize(temperature = mean(temperature))
```


### 60% Elevation: 2000 dataset, 2010-2020, spring (130) down/summer (215) up 0.8 C

```{r}
datats %>%
  bucketize("elev", "m", 10) %>%
  filter(Bucket == 60) %>%
  group_by(StartYr, Year, day) %>%
  summarize(temperature = mean(temperature)) %>%
  group_by(StartYr, day) %>%
  arrange(Year) %>%
  mutate(temperature = zoo::rollmean(temperature, 5, fill=NA, align="right")) %>%
  ungroup() %>%
  filter(StartYr == 2000, Year %in% c(2010, 2020), day %in% c(130, 215)) %>%
  group_by(Year, day) %>%
  summarize(temperature = mean(temperature))
```

## Spatial Trends

Based on strength/weakness of spsu and aw anomalies, divergences from sinusoid
should be obvious for western Montana and northern Maine, but negligible for Florida.
Make plots of actual cycle vs sinusoid for:

- Western Montana = -115 to -110; 45+
- Northern Maine = -70+; 45+
- Florida = -87+; 30-

```{r}
frn <- function(lons, lats, name, dat=geo) {
  filter(dat,
         lon >= lons[1], lon <= lons[2],
         lat >= lats[1], lat <= lats[2]) %>%
    mutate(Region = name)
}

regions <- rbind(
  frn(c(-115, -110), c(45, 99), "Western Montana"),
  frn(c(-70, 0), c(45, 99), "Northern Maine"),
  frn(c(-87, 0), c(0, 30), "Florida")
)
```

```{r}
rfit <- regions %>%
  group_by(Region, id, day) %>%
  summarize(temperature = mean(temperature)) %>%
  group_by(Region) %>%
  mutate(
    Cos = cos((day - 210) * 2 * pi / 365),
    Fit = lm(temperature ~ Cos)$fitted.values,
    N = length(unique(id))
  ) %>%
  group_by(Region, day) %>%
  mutate(Mean = mean(temperature)) %>%
  ungroup() %>%
  mutate(Region = paste0(Region, " (n=", N, ")"))
```

```{r}
plt <- ggplot(rfit, aes(day)) +
  geom_point(aes(y=temperature, color="Observed"), size=2, alpha=0.1,
             stroke=NA) +
  geom_line(aes(y=Fit, color="Sinusoid"), size=4) +
  geom_line(aes(y=Mean, color="Mean"), size=2) +
  facet_wrap(~Region, nrow=1, scales="free_y") +
  scale_color_viridis_d() +
  theme_std() +
  theme(legend.position = "bottom") +
  labs(
    x="Day of Year",
    y="Region Mean Temperature",
    color=NULL
  )
write.img(plt, "Figures/GeoValidation", height=3)
```

# Temporal Cross-Validation

```{r}
fitted <- data %>%
  sinu.resfit %>%
  drop_na() %>%
  bucketize("elev", "m", 10) %>%
  group_by(id) %>%
  group_modify(~left_join(.x, make.sints(fit.sins(.x)), by="day"))
```

```{r}
cv <- data %>%
  mutate(Year = as.integer(format(Date, "%Y"))) %>%
  group_by(id) %>%
  group_modify(~{
    yrs <- unique(.x$Year)
    cvdat <- left_join(.x,
                      tibble(Year = yrs,
                             Index = sample(1:5, length(yrs), replace=TRUE)),
                      by="Year")
    map_dfr(1:5,
            function(ix) {
              tr <- filter(cvdat, Index != ix)
              te <- filter(cvdat, Index == ix) %>% select(-Index)
              model <- fit.sins(tr)
              days <- 1:366
              modout <- tibble(day=days,
                               Sinusoid=model$Intercept + model$Amplitude * cos((days - 210) * 2 * pi / 365),
                               Anomaly=make.sints(model)$actemp)
              left_join(te,
                        modout,
                        by="day")
            })
  })
```
```{r}
perf <- cv %>%
  group_by(id) %>%
  summarize(
    across(c(da, elev), first),
    across(c(Sinusoid, Anomaly),
    ~cor(.x, temperature)^2
  )) %>%
  mutate(Delta = Anomaly - Sinusoid)

perf %>%
  select(Sinusoid, Anomaly, Delta) %>%
  summary
```

## CV Performance Trends

```{r}
cor(perf$elev, perf$Delta)
cor(perf$da, perf$Delta)
perf %>%
  group_by(Elev500 = 500 * floor(elev/500)) %>%
  summarize(
    across(Sinusoid:Delta,
           median)
  )
```


# Predicting Seasonality Functions

From Isaak et al. 2020: elevation, latitude, riparian canopy density,
reach slope, precipitation, lake prevalence, and dam height

Summarized through PCA.  However, they were using classification categories,
not continuous values.


# Misc

## Gage Counts and Data Coverage

```{r}
plt <- ggplot(tg, aes(x=tslen / 365)) +
  stat_ecdf(size=4) +
  scale_y_continuous(sec.axis = sec_axis(~.*nrow(tg), name="Gage Count")) +
  theme_std() +
  labs(x="Gage Timeseries Length (years)",
       y="Cumulative Probability")
write.img(plt, "Figures/GageECDF", width=4)
```


## Plot of Data Coverage

```{r}
plt <- plot.eco(TRUE) +
  scale_fill_viridis_d(option="magma", end=0.5) +
  geom_point(aes(lon, lat, color=tslen/365, size=tslen/365), data=tg) +
  scale_color_viridis_c(begin=0.5) +
  theme_std() +
  labs(
    x="Longitude",
    y="Latitude",
    color="Timeseries Length (years)",
    size="Timeseries Length (years)",
    fill="EPA Level I Ecoregion"
  )
write.img(plt, "Figures/GageCovMap")
```

## Arkansas River Demo

This is the Arkansas River at Granite, CO, in between Leadville and Buena Vista
at about 9,000 ft.

```{r}
gn <- "07086000"
data <- read_csv("Data/GageData.csv",
                 col_types = "cdcdDid") %>%
  drop_na %>%
  filter(id == gn)
```

```{r}
data <- data %>%
  group_by(day) %>%
  mutate(tmean = mean(temperature)) %>%
  ungroup() %>%
  mutate(tsin = lm(tmean ~ cos((day - 210) * 2 * pi / 365))$fitted.values)
data <- left_join(data,
                  make.sints(fit.sins(data)),
                  by="day")
```
```{r}
plt <- ggplot(data, aes(day)) +
  geom_point(aes(y=temperature, color="Observations"),
             alpha=0.1, stroke=NA, size=2) +
  geom_line(aes(y=actemp, color="Three-Sine"), size=2) +
  geom_line(aes(y=tsin, color="Sine"), size=2) +
  scale_color_viridis_d(end=0.8) +
  theme_std() +
  labs(
    x="Day of Year",
    y="Gage Temperature (C)\n(Arkansas River at Granite, CO)",
    color=NULL
  )
write.img(plt, "Figures/ArkansasDemo", width=4)
```

## Conceptual Diagram

```{r}
# sysin <- tibble(
#   Intercept = 12,
#   Amplitude = 10,
#   FallDay = 330,
#   WinterDay = 40,
#   FallWinter = 3,
#   SpringDay = 150,
#   SummerDay = 230,
#   SpringSummer = 2
# )
sysin <- fit.sins(data)
ts <- make.sints(sysin) %>%
  rename(Value = actemp) %>%
  mutate(Component = "Overall Function (Sum)",
         Which = "All")
days <- ts$day
sinpt <-
  tibble(
    day = days,
    Value = sysin$Intercept + sysin$Amplitude * cos((day - 210) * 2 * pi / 365),
    Component = "Main Sine",
    Which = "Components"
  )
anoms <- mutate(ts,
               Value = Value - sinpt$Value,
               Which = "Components")
awin <- filter(anoms, day >= 250 | day <= 95) %>%
  mutate(Component = "Autumn/Winter Anomaly",
         Value = ifelse(abs(Value) == 0, NA, Value))
ssu <- filter(anoms, day > 95, day < 250, abs(Value) > 0) %>%
  mutate(Component = "Spring/Summer Anomaly")

inpt <- rbind(ts, sinpt, awin, ssu)
```

```{r}
plt <- ggplot(inpt, aes(day, Value, color=Component)) +
  geom_line(size=2) +
  facet_wrap(~Which, nrow=2) +
  scale_color_viridis_d(end=0.8) +
  theme_std() +
  labs(
    x="Day of Year",
    y="Mean Stream Temperature",
    color="Function Component"
  )
write.img(plt, "Figures/ConceptualDiagram")
```


## Coverage

```{r}
data <- left_join(data, tg, by=c("id"="site_no"))
```
```{r}
unq <- data %>%
  group_by(id) %>%
  slice_head(n=1) %>%
  ungroup()
```


```{r}
c(
  "East" = nrow(filter(unq, lon >= -100)) / nrow(unq),
  "Alaska" = nrow(filter(unq, lat > 50)),
  "Hawaii" = nrow(filter(unq, lon < -140, lat < 30))
)
```

```{r}
ecodata <- left_join(data, read_csv("Data/ecoregions.csv") %>%
                       select(id, ecoregion), by="id")
```
```{r}
ecodata %>%
  group_by(ecoregion) %>%
  summarize(
    N = length(unique(id))
  ) %>%
  ungroup() %>%
  mutate(Frac = N / sum(N))
```



